---
author: $\overset{\mathrm{martin.o.berild@ntnu.no}}{10014}$ \and
        $\overset{\mathrm{yaolin.ge@ntnu.no}}{10026}$
date: \today
title: "Project 3 - TMA4315"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
defaultW <- getOption("warn") 
options(warn = -1) 
```

## Problem 1 

#### a)
It can be shown as 
Assume data $(y_{ij}, x_{ij}^T)$ for $i = 1, 2, ..., m, j = 1, 2, ..., n_i$


$$
y_{ij} = x_{ij}^T\beta + u_{ij}^T\gamma_i + \epsilon_{ij}
$$
By applying the random intercept slope model

$$
y_{ij} = \begin{bmatrix} 1 & x_{ij}\end{bmatrix} \begin{bmatrix} \beta_0 \\ \beta_1\end{bmatrix} + \begin{bmatrix} 1 & x_{ij}\end{bmatrix} \begin{bmatrix} \gamma_{0, i} \\ \gamma_{1, i}\end{bmatrix} + \epsilon_{ij}
$$
Letting

$$
y_i = \begin{bmatrix} y_{i1} \\ y_{i2} \\ $\vdots$ \\ y_{in_i} \end{bmatrix}, \ \ x_i = \begin{bmatrix} x_{i1}^T \\ x_{i2}^T \\ \vdots \\ x_{in_i}^T \end{bmatrix}, \ \ U_i = \begin{bmatrix} u_{i1}^T \\ u_{i2}^T \\ \vdots \\ u_{in_i}^T \end{bmatrix}, \ \ \epsilon_i = \begin{bmatrix} \epsilon_{i1} \\ \epsilon_{i2} \\ $\vdots$ \\ \epsilon_{in_i} \end{bmatrix}
$$



The the model is (for cluster $i = 1, 2, \cdots, m$):
$$
y_i = X_i\beta + U_i\gamma_i + \epsilon_i, \ \ \gamma_i \sim \mathcal{N(0, Q)}
$$


The marginal model is then

$$
y_i \sim \mathcal{N}(X_i \beta,\ \  U_iQU_I^T + \sigma^2I_{n_i})
$$

Conditional model is
$$

y_i|\gamma_i \sim \mathcal{N}(X_i\beta + U_i\gamma_i, \ \ \sigma^2I_{n_i})
$$

<!-- Letting -->
<!-- $$ -->

<!-- y = \begin{bmatrix} y_{1} \\ y_{2} \\ $\vdots$ \\ y_{m} \end{bmatrix}, \ \ X = \begin{bmatrix} X_{1}^T \\ X_{2}^T \\ $\vdots$ \\ X_{m}^T \end{bmatrix}, \ \ \gamma = \begin{bmatrix} \gamma_{1}^T \\ \gamma_{2}^T \\ $\vdots$ \\ \gamma_{m}^T \end{bmatrix}, \ \ \epsilon = \begin{bmatrix} \epsilon_{1} \\ \epsilon_{2} \\ $\vdots$ \\ \epsilon_{m} \end{bmatrix} -->

<!-- $$ -->

<!-- Then the whole model is -->

<!-- $$ -->
<!-- y = X\beta + U\gamma + \epsilon, \ \ \epsilon \sim \mathcal{N}(0, \sigma^2 I_{n_i}) -->
<!-- $$ -->

We are interested in computing the maximum likelihood and restricted maximum likelihood estimates of the parameters in the linear mixed model 

$$
y_{ij} = \beta_0 + \beta_1x_{ij} + \gamma_{0,i} + \gamma_{1,i}x_{ij} + \epsilon_{ij},
$$
where $\boldsymbol{\gamma}_i = (\gamma_{0,i},\gamma_{1,i})$ are iid binomially distributed with zero mean and covariance matrix

$$
\begin{bmatrix}
\tau_0^2 & \tau_{01} \\
\tau_{01} & \tau_{1}^2
\end{bmatrix},
$$
and $\epsilon_{ij}$ are idd Guassian distributed with zero mean and variance $\sigma^2$ for $i = 1,\dots,m$ and $j=1,\dots,n$. Thus, we are interested in obtaining the estimates of $(\beta_0,\beta_1,\tau_0^2,\tau_1^2,\tau_{01},\sigma^2)$. 


## Problem 2

Next, we are interested in modelling the 2018 results of the Norwegian elite football league using a generalized linear mixed model. First we load the data and display the contents.

```{r dataload}
long <- read.csv("https://www.math.ntnu.no/emner/TMA4315/2020h/eliteserie.csv", colClasses = c("factor","factor","factor","numeric"))
head(long)
```



```{r endup, include=FALSE}
options(warn = defaultW)
```